# Tidal analysis of Tuktoyaktuk sea level

**Conclusion.** The results match those in Foreman's manual on tidal analysis,
which is also the source of the data used here.

```{r}
library(oce)
data(sealevelTuktoyaktuk)
```

Tidal model fitted with tidem()
```{r}
tide <- tidem(sealevelTuktoyaktuk)
summary(tide)
```
Extract data for plotting
```{r}
time <- sealevelTuktoyaktuk[["time"]]
eta <- sealevelTuktoyaktuk[["elevation"]]
```

Plot the tidal model, created by a integrating component amplitudes over
frequency.  Peak of graph says the total tidal signal is approximately 1.7m.
```{r tidemodel,fig.path="figure/tidal_analysis",fig.cap="**Tidal model.**"}
plot(tide)
```

Plot observed sea level.  Note the data gaps, which pose no problem with
regression-based tidal analysis but would be problematic for spectral analysis.
```{r tide,fig.path="figure/tidal_analysis",,fig.cap="**Observed sea level.**",fig.width=9}
oce.plot.ts(time, eta, type='l', main="Sea level [m]", grid=TRUE) 
```

Plot the tidal residual.  (Do some audience members have ideas on what causes
the variability?)
```{r residual,fig.path="figure/tidal_analysis",,fig.cap="**Tidal residual.**",fig.width=9}
oce.plot.ts(time, eta-predict(tide), type='l', main="Residual sea level [m]", grid=TRUE)
```

The example shows:

* using built in datasets
* using tidem()
* using predict(), a *generic* function (a very important idea)
* using summary(), again a generic
* plotting time series with extra features like the time range

Ideas for making it live:

* turn off the time range
* ``table(is.na(eta))`` produces a very useful result with little effort
* ``fivenum(eta)`` again is useful
* ``hist(eta)`` , ``boxplot(eta)`` and similar plots may be informative

