# Outlier detection in a TS diagram

**Conclusion.** R statistical tools are useful in identifying TS outliers.

```{r}
library(oce)
data(ctd)
```

Get an idea on the data
```{r ctd,fig.path="figure/TS_outlier",fig.cap="**TS diagram.  Note the scatter.**"}
plotTS(ctd)
```

Illustrate the ``[[]]`` method
```{r}
S <- ctd[["salinity"]]
T <- ctd[["temperature"]]
```
NOTE: *never* use the TRUE=T notation!

Get audience opinion on df, from eyeballing graph
```{r}
df <- 3
m <- smooth.spline(S, T, df=df)
lines(predict(m)) # DISCUSS: it looks for "x" and "y" in the data
```

S is not monotonic so this idea is nuts...go to density-spice space
```{r sigthe_spice,fig.cap="**sigma_theta - spice diagram.**"}
sigthe <- ctd[["sigmaTheta"]]
spice <- ctd[["spice"]]
plot(sigthe, spice)
```

New name for this spline -- not a good name but one easy to type (practical).
```{r}
mm <- smooth.spline(sigthe, spice, df=df)
lines(predict(mm))
```

Probably want larger df ... can play a bit.
```{r}
df <- 10
mm <- smooth.spline(sigthe, spice, df=df)
lines(predict(mm), col='red')
```

The predictions are at *data locations*; optionally show them how to get a
smooth curve but for the present purpose we actually *want* the predictions at
the data locations.

Deviation histogram
```{r histogram,fig.path="figure/TS_outlier",fig.cap="**Histogram of spice deviation from spline.**"}
dev <- spice - predict(mm)$y
hist(dev)
```

Play with breaks; can show result of too many; maybe also show boxplot.
```{r boxplot,fig.path="figure/TS_outlier",fig.cap="**Boxplot of spice deviation from spline.**"}
boxplot(dev)
```

Note the outliers ... but we will make our own definition.

Q to audience: anyone recognize this number?
```{r}
qnorm(0.975)
```

Note that we can do the "t" distribution as well!
```{r}
qt(0.975, df=9999)
```
note asymptote.  Nnow do for our
```{r}
qt(0.975, df=length(spice))
```

OK now for our method. Construct "z" values and identify those that seem
unusual at 95% level.
```{r}
z <- (dev - mean(dev)) / sd(dev)
bad <- abs(z) > qt(0.975, df=length(spice))
table(bad) 
```
Plot to see if we like the results
```{r sigthe_spice_summary,fig.path="figure/TS_outlier",fig.cap="**Data with (blue) spline; red indicates outliers.**"}
plot(sigthe, spice, col=ifelse(bad, "red", "black"), pch=20)
lines(predict(mm), col='blue')
```

Finally, a summary in the more familiar TS space. 
```{r summary,fig.path="figure/TS_outlier",fig.cap="**Summary diagram, with red for anomalous points.**"}
plotTS(ctd)
theta <- ctd[["potential temperature"]]
points(S[bad], theta[bad], col="red", pch=20, cex=2)
```
