**sandbox/dk/03: read oddly-formatted (SOI) data**

Sometimes data sets are simple, consisting of known binary or text formats, but
sometimes we have to deal with odd formats that are one of a kind.  The SOI
(Southern Oscillation Index) dataset at 
[http://www.esrl.noaa.gov/psd/data/correlation/soi.data](http://www.esrl.noaa.gov/psd/data/correlation/soi.data) is a case in point  The first few lines are
```
        1948        2014
1948 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99
1949 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99
1950 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99
1951   2.50   1.50  -0.20  -0.50  -1.10   0.30  -1.70  -0.40  -1.80  -1.60  -1.30  -1.20
1952  -1.50  -1.00   0.90  -0.40   1.20   1.20   0.80   0.10  -0.40   0.60   0.00  -2.00
```
Perhaps the format is defined somewhere (but more likely not), so what to do?

The first-order options are (a) give up until proper documentation is found or
(b) make some reasonable guesses.  So, what's reasonable?

* The first line is probably a time range.
* The other lines start with a number that could be a year, and the 12
  following numbers may be monthly values (at start of month? middle?)
* The value ``-99.99`` is likely a missing-value code.

This suggests an easy way to read the data:
```{r}
d <- read.table("http://www.esrl.noaa.gov/psd/data/correlation/soi.data", skip=1, header=FALSE)
```
but there is an error.  Looking at the end of the file we see
```
2013  -0.10  -0.40   2.50   0.40   1.30   2.00   1.30   0.30   0.50  -0.10   1.20   0.10
2014   2.40   0.10  -1.50   1.30 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99 -99.99
  -99.99
  SOI Index from CPC
 http://www.esrl.noaa.gov/psd/data/climateindices/list/ for info
```
so this file evidently has a trailer as well as a header.  R is great with
headers but trailers are harder to deal with.  A human might look at the line
with a single number and think, "ahah that is telling us the missing-value
code", and similarly would take some meaning in the other two lines.

What to do?  

**The cautious approach.** Verify these guesses about the data, by searching
for documentation on the format.  But we may never actually find the format,
and it could change at any time...

**The quick and dirty approach.**  Download the data into a file, make a copy
of the file and edit the header and trailer out of the new file, then work with
that.

**The somewhat more elegant approach.** Read the file with R, and apply some
logic to find the data lines.  One approach might be to accept only lines with
13 tokens, the first of which is a number that could be a year.
```{r}
d <- readLines("http://www.esrl.noaa.gov/psd/data/correlation/soi.data")
dlen <- unlist(lapply(seq_along(d), function(i) nchar(d[i])))
```
(This is a quirk of the author -- calling temporary variables ``d`` to save typing.)

Now, examine the lengths
```{r}
dlen
```
This suggests something one might notice looking at the website: the data lines are all of 
the same length.  So we select just lines of that length
```{r}
d2 <- d[dlen == 88]
```
(Notice we keep ``d``, just in case ... always a good idea in interactive work.)

Now we can ``scan()`` to get the data
```{r}
d3 <- scan(text=d2)
```
To clean up, we extract the years and make ``-99.99`` into a missing value.
```{r}
d4 <- d3[d3 < 1900]
SOI <- d4[d4 > -99]
```
Notice the name switch from ``d`` (for some sort of temporary data) to ``SOI`` (for quantity
of interest).

It remains to construct a time vector
```{r}
year <- seq(from=d3[1], by=1/12, length.out=length(SOI))
```

Let's plot the results
```{r soiplot,fig.cap="**Southern Oscillation Index (inferred here)**"}
plot(year, SOI, type='l')
```

Postscript: serious work would, of course, involve some reading.  Just what the
heck is this index?  There are several available online, with different
definitions... for example, the ``soi`` dataset in the ``oce`` package has a
similar pattern but different values.

```{r soiplot2,fig.cap="**Southern Oscillation Index (in oce package)**"}
library(oce)
data(soi)
plot(soi$year, soi$index, type='l', xlim=range(year))
```
