**sandbox/dk/04: imagep() with colormap**

Will use salinity from an Argo drifter (dataset included with oce).
```{r}
library(oce)
data(drifter)
```

Get an idea of the data contents with ``summary()``, a "generic" function whose
action depends on the class of the object it is examining.
```{r}
summary(drifter)
```

Extract pressure, time, and salinity.  Note that pressure is a matrix, and that
salinity is in a matrix that we transpose for plotting.
```{r}
p <- drifter[["pressure"]][,1] # note that p is a matrix
t <- drifter[["time"]]
S <- t(drifter[["salinity"]]) # note the transpose
```

For more details on salinity, try a histogram (the ``xlim`` came from fiddling
around)
```{r 04-histogram,fig.cap="Trimmed salinity histogram"}
hist(S, xlim=c(33.5, 36), breaks=200)
```

OK, time to make a salinity section.  Construct a colourmap based on this
```{r}
cmS <- colormap(zlim=c(33.5, 36), col=oceColorsJet)
```
and then make an image graph (with a contour for fun)
```{r 04-section,fig.cap="Salinity section in time-pressure space"}
imagep(t, p, S, colormap=cmS, ylim=rev(range(p)))
contour(t, p, S, level=35, add=TRUE, drawlabels=FALSE) # check a colour
```

Anyone think salinity was changing over time? To check, we can make a
time-coded trajectory graph.
```{r 04-map,fig.cap="Time-coded drifter trajectory"}
year <- 2005 + (as.numeric(t) - as.numeric(as.POSIXct("2005-01-01"))) / 365 / 86400
cmt <- colormap(year, col=oceColorsJet)
drawPalette(colormap=cmt, zlab="Year")
plot(drifter, pch=21, bg=cmt$col, mar=par('mar')) # note the mar trick
```
