# Base graphics example

When it comes to making plots in R, there are a wide variety of options. Popular packages that extend the base plotting capabilities of R include the [`ggplot2`](http://cran.r-project.org/web/packages/ggplot2/index.html) and [`lattice`](http://cran.r-project.org/web/packages/lattice/index.html) packages.

The `oce` package however, almost exclusively uses the base graphics plotting functions. Thus to understand how to tweak, manipulate, and add to plots create through `oce` functions it is a good idea to have an idea how to interact with R's basic graphing system.

## The Basics

Let's start with a basic plot, created from the `adp` dataset in `oce`. For simplicity, we'll extract some of the variables from the adp object, to make the plotting commands cleaner.

```{r}
library(oce)
data(adp)
time <- adp[['time']]
pressure <- adp[['pressure']]
temperature <- adp[['temperature']]
```

We can plot the variables using the `plot()` command. Remember that R and oce use objects and "generic" functions, so if we do something like `plot(adp)`, we are in fact calling the function `plot.adp()` which has different behavior from the default `plot.default()` function (see e.g. `?plot.adp`).

```{r}
plot(pressure)
```
makes a plot of the pressure vector against the indices of the vector -- this is the default behavior when only the `x` argument is given. Note that the default plot type is to use points. This can be changed using the `type` argument (see `?plot.default` for details):
```{r}
par(mfrow=c(2,2)) # 2 by 2 grid
plot(pressure, type='p') # default
plot(pressure, type='l') # lines
plot(pressure, type='b') # both
plot(pressure, type='o') # overplotted both
```

We can plot temperature as a function of pressure:
```{r}
par(mfrow=c(1,1))
plot(pressure, temperature)
```
or we can use the `~` formula notation to plot as the form `plot(y ~ x)`
```{r}
plot(temperature ~ pressure)
```


Notice that every time you use the `plot()` command, it draws a new plot. This is what is known as a "high-level" plotting function. To add further things to the plot, such as more points, lines, text, legends, etc, requires the use of a series of so-called "low-level" plotting commands.

Let's add some points perturbed from the original by random noise, and use some different "plotting characters", as specified by the `pch` argument:
```{r}
plot(pressure, temperature, pch=21, bg='grey')
pnoise <- rnorm(pressure, sd=sd(pressure)/10)    # first argument specifies desired length
tnoise <- rnorm(length(temperature), sd=sd(temperature)/10)   # can be a vector or a single number
points(pressure + pnoise, temperature + tnoise, pch=22, bg='red', cex=0.5)
```

How about we add a line, such as a least squares fit of T(p) (i.e. a "linear model"):
```{r}
plot(pressure, temperature, pch=21, bg='grey')
grid()   # add a grid
pnoise <- rnorm(pressure, sd=sd(pressure)/10)    # first argument specifies desired length
tnoise <- rnorm(length(temperature), sd=sd(temperature)/10)   # can be a vector or a single number
points(pressure + pnoise, temperature + tnoise, pch=22, bg='red', cex=0.5)
m <- lm(temperature ~ pressure)
press.pred <- seq(min(pressure), max(pressure), 0.01)
temp.pred <- predict(m, list(pressure=press.pred))
lines(press.pred, temp.pred, lwd=10, col='pink')
abline(m)      # can add a straight line directly using the lm object
```


## Tweaking the aesthetics

Notice that the default margins leave lots of room for readability, but may be too large for most practical purposes. To adjust all the different aesthetic aspects of the plot, we make use of the myriad of different "parameters" -- to see all of them do `?par` or type `par()` at the command line. Note that we already used the `mfrow` parameter to set the layout of subplots.

As an example here, we will adjust both the "margins" (through the `mar` parameter) and the "margin positions" (through the `mgp` parameter). We need to call `par()` to set the various parameters *before* doing any plotting. In general, a call to `par()` creates a new device instance, which will result in an empty plot, so you don't want to try it halfway through a series of low-level plotting commands.

The current par values can be queried using the `par()` function -- e.g. if we want to know what the current settings are for `mar`, we do:
```{r}
par('mar')
```
These numbers can be interpreted as the number of "lines" around the outside of the plot, starting with the bottom, then the left side, then the top, then the right side. Similarly we can query the `mgp`
```{r}
par('mgp')
```
These 3 numbers represent the line that axis label is printed on, the line the tick labels are printed on, and the line that the axis itself is drawn on.

Let's try changing them to make a tighter plot:
```{r}
par(mar=c(3, 3, 1, 1), mgp=c(2, 0.75, 0))
plot(pressure, temperature, pch=21, bg='grey')
grid()  
points(pressure + pnoise, temperature + tnoise, pch=22, bg='red', cex=0.5)
```
Notice that we've reduced the space around the plot box, and have tightened the labeling of the axes. If we wanted, we could move the axis position *away* from the zero position:
```{r}
par(mar=c(3, 3, 1, 1), mgp=c(2, 0.75, 0.25))
plot(pressure, temperature, pch=21, bg='grey')
grid()  
points(pressure + pnoise, temperature + tnoise, pch=22, bg='red', cex=0.5)
```

## Legends





To include:

* high-level plotting commands (e.g. `plot()`)

* low-level plotting commands (e.g. `points()`, `lines()`, ... )

* multi-figure plots (e.g. `mfrow/mfcol` and `layout()`)

* graphing parameters, i.e. `par`

    * `mar`, `mgp`, ...

	* axes styles (e.g. `xaxs`)

* adding text (e.g. `text()`, and `mtext()`)

* legends

* point out commonly used missing (or inflexible) functions, such as image plots, filled contours, palettes and colormaps.
